---
title: "Lab 3A"
author: "Vista Sohrab"
date: "9/7/2020"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Data Carpentry - Starting with data

## Presentation of Survey Data
### Downloading the animal survey data
```{r download data, eval = FALSE}
download.file(url = "https://ndownloader.figshare.com/files/2292169", destfile = "data/portal_data_joined.csv")

```

### Loading the animal height and weight data and viewing the resulting dataframe:
```{r read input}
animal_survey <- read.csv("data/portal_data_joined.csv")

head(animal_survey)

str(animal_survey)
```

### Inspection of the animal survey dataframe
```{r inspect df}
# Dataframe size, number of rows and columns
dim(animal_survey)
nrow(animal_survey)
ncol(animal_survey)

#Dataframe content
head(animal_survey)
tail(animal_survey)

#Dataframe column and row names
names(animal_survey)
head(rownames(animal_survey))

#Dataframe structure and summary
str(animal_survey)
summary(animal_survey)
```

## Challenge 1

* The class of the animal_survey object is a dataframe.
* The dataframe has 34786 rows and 13 columns.
* The species_id column contains 48 factors; therefore, there are 48 different species recorded in the survey.

## Indexing and Subsetting Dataframes

### Subsetting via indices
```{r subset with indices}
#Extract value in row 1, column 1
animal_survey[1,1]

#Extract value in row 6, column 1
animal_survey[6,1]

#Extract first column of dataframe in vector form
head(animal_survey[, 1])

#Extract first column of dataframe as data.frame
head(animal_survey[1])

#Extract first three elements of the 7th column
animal_survey[1:3, 7]

#Extract third row as data.frame
animal_survey[3, ]

#Statement equivalent to head of a dataframe by obtaining first 6 rows
animal_survey[1:6, ]

#Exclude the first column of dataframe
head(animal_survey[, -1])

#Exclude row 7 to end of dataframe, equivalent to head of dataframe
animal_survey[-(7:nrow(animal_survey)),]
```

### Subsetting via column names

```{r subset with col names}
#Extract species_id column as data.frame object
head(animal_survey["species_id"])

#Extract species_id column as vector
head(animal_survey[, "species_id"])
head(animal_survey[["species_id"]])
head(animal_survey$species_id)

```

## Challenge 2

```{r}
survey_row200 <- animal_survey[200, ]
survey_row200
```


```{r}
survey_row34786 <- animal_survey[34786, ]
survey_row34786

tail(animal_survey)

survey_lastRow <- animal_survey[nrow(animal_survey), ]
survey_lastRow

```


```{r}
survey_middle <- animal_survey[nrow(animal_survey)/2 ,]
survey_middle
```

```{r}
alt_head_survey <- animal_survey[-(7:nrow(animal_survey)),]
alt_head_survey
```

## Factors

```{r}
sex <- factor(c("male","female","female", "male"))
levels(sex)
nlevels(sex)

#current order of levels
sex

#change order of levels
sex <- factor(sex, levels = c("male", "female"))
sex

```

### Converting Factors

```{r}
#converting factors into strings
as.character(sex)
```


```{r}

#creating year_fct vector
year_fct <- factor(c(1990, 1983, 1977, 1998, 1990))

#incorrect, returning index values of factor not levels
as.numeric(year_fct)

#correct ways of converting factors into numerics
as.numeric(as.character(year_fct))
as.numeric(levels(year_fct)[year_fct])
```

### Renaming factors

```{r}
# bar plot of number of males and females of the animal survey 
plot(as.factor(animal_survey$sex))
```


```{r}
sex <- factor(animal_survey$sex)
#modifying level names
head(sex)
levels(sex)
levels(sex)[1] <- "undetermined"
levels(sex)
head(sex)
```

## Challenge 3
```{r}
levels(sex)[2:3] <- c("female", "male")
head(sex)
```


```{r}
sex <- factor(sex, levels=c("female", "male", "undetermined"))
plot(sex)
```

### Using stringsAsFactors=FALSE
```{r}
animal_survey_factor <- read.csv("data/portal_data_joined.csv", stringsAsFactors = TRUE)

str(animal_survey_factor)

animal_survey_char <- read.csv("data/portal_data_joined.csv", stringsAsFactors = FALSE)

str(animal_survey_char)
```


```{r}
str(animal_survey_char$plot_type)
#convert plot_type column into factor
animal_survey_char$plot_type <- factor(animal_survey_char$plot_type)
str(animal_survey_char$plot_type)
```

## Challenge 4

The errors present in the original code were lack of double quotes around strings when concatenating animal values, the weight column missing a comma separating one value from another, as well the feel column missing an entry for cat causing a differing number of rows error since all columns in a single dataframe should have the same length. 
```{r}
animal_data <- data.frame(animal= c("dog", "cat", "sea cucumber", "sea urchin"),
                          feel= c("furry", "furry", "squishy","spiny"),
                          weight = c(45, 8, 1.1, 0.8))
```

I would expect country, climate, temperature, and northen hemisphere columns to be treated as factors, since the data.frame function converts strings into factors as its default and these four columns contain at least one element enclosed in double quotes implying a string.
```{r}
country_climate <- data.frame(
       country = c("Canada", "Panama", "South Africa", "Australia"),
       climate = c("cold", "hot", "temperate", "hot/temperate"),
       temperature = c(10, 30, 18, "15"),
       northern_hemisphere = c(TRUE, TRUE, FALSE, "FALSE"),
       has_kangaroo = c(FALSE, FALSE, FALSE, 1)
       )
str(country_climate)
```


If stringsAsFactors is set to false, then the country, climate, temperature, northern_hemisphere will be characters due to the presence of double quotes in at least one of their entries.


```{r}
country_climate_char <- data.frame(
       country = c("Canada", "Panama", "South Africa", "Australia"),
       climate = c("cold", "hot", "temperate", "hot/temperate"),
       temperature = c(10, 30, 18, "15"),
       northern_hemisphere = c(TRUE, TRUE, FALSE, "FALSE"),
       has_kangaroo = c(FALSE, FALSE, FALSE, 1),
       stringsAsFactors = FALSE)

str(country_climate_char)
```

To achieve the intended data types, the removal of double quotes from 15 in temperature and FALSE in northern_hemisphere will lead to temperature becoming numeric and northern_hemisphere becoming logical. Additionally, changing the 1 in has_kangaroo to True will also result in has_kangraoo becoming logical.
```{r}
country_climate <- data.frame(
       country = c("Canada", "Panama", "South Africa", "Australia"),
       climate = c("cold", "hot", "temperate", "hot/temperate"),
       temperature = c(10, 30, 18, 15),
       northern_hemisphere = c(TRUE, TRUE, FALSE, FALSE),
       has_kangaroo = c(FALSE, FALSE, FALSE, TRUE),
       stringsAsFactors = FALSE)

str(country_climate)
```

## Formatting Dates

```{r warning=FALSE}
#import lubridate package
library(lubridate)
```

```{r}
# observe numeric month, day and year columns in animal survey data
str(animal_survey)
```


```{r}
this_date <-ymd("2020-09-07") 
str(this_date)

#alternative way with paste
alt_this_date <- ymd(paste("2020","9", "7",sep="-"))
str(alt_this_date)
```

Apply ymd to animal_surveys dataframe:

```{r}
animal_survey_dates <- paste(animal_survey$year, animal_survey$month, animal_survey$day, sep = "-")
head(animal_survey_dates)

```


```{r}
animal_survey$date <- ymd(paste(animal_survey$year, animal_survey$month, animal_survey$day, sep = "-"))
str(animal_survey)
summary(animal_survey$date)
missing_dates <- animal_survey[is.na(animal_survey$date), c("year", "month", "day")]
head(missing_dates)
```

There are 129 dates that failed to parse, since these dates are in incorrect format as such dates do not exist in the calendar. From viewing the first few rows of missing_dates, September and April are months with 30 days; therefore, there has been an error during data entry. These dates can be changed to the 30th of the month in order to continue using these instances for data analysis or they can be dropped from the data frame depending on the context. 