---
title: "Lab 3B"
author: "Vista Sohrab"
date: "9/7/2020"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r warning=FALSE}
library(tidyverse)
```

# Data Manipulation using dplyr and tidyr

### Importing data and inspecting
```{r import data}
animal_survey <- read_csv("data/portal_data_joined.csv")
str(animal_survey)
```

### Selecting Columns and Filtering Rows

```{r select}
# selecting plot_id, species_id and weight columns
select(animal_survey, plot_id, species_id, weight)

# excluding record_id and species_id 
select(animal_survey, -record_id, -species_id)

#choosing rows based on specific criterion
filter(animal_survey, year == 1995)
```

## Pipes

Intermediate steps:
```{r intermediate steps}
animal_survey2 <- filter(animal_survey, weight < 5)
animal_survey_sml <- select(animal_survey2, species_id, sex, weight)
animal_survey_sml
```

Nest functions:
```{r nested}
animal_survey_sml_nested <- select(filter(animal_survey, weight < 5), species_id, sex, weight)
animal_survey_sml_nested
```

Pipes:
```{r}
animal_survey %>% 
  filter(weight < 5) %>% 
  select(species_id, sex, weight)
```

Create subsetted dataframe:
```{r}
animal_survey_sml_piped <- animal_survey %>% 
  filter(weight < 5) %>% 
  select(species_id, sex, weight)
animal_survey_sml_piped
```

## Challenge 1

```{r}
animal_survey_pre95 <- animal_survey %>% 
  filter(year < 1995) %>% 
  select(year, sex, weight)

animal_survey_pre95
```

### Mutate

Create new column of weight in kilograms:

```{r}
animal_survey %>% 
  mutate(weight_kg = weight/1000) %>% 
  head()

```

Create column of weight in kilograms and weight in pounds:
```{r}
animal_survey %>% 
  mutate(weight_kg = weight/1000,
         weight_lb = weight_kg * 2.2) %>% 
  head()
```

Filter out NAs:
```{r}
animal_survey %>%
  filter(!is.na(weight)) %>%
  mutate(weight_kg = weight / 1000) %>%
  head()
```

## Challenge 2

```{r}
animal_survey_hindfoot_cm <- animal_survey %>% 
  filter(! is.na(hindfoot_length)) %>% 
  mutate(hindfoot_cm = hindfoot_length / 10) %>% 
  select(species_id, hindfoot_cm) %>% 
  filter(hindfoot_cm < 3)

animal_survey_hindfoot_cm
```

### Split-apply-combine data analysis

```{r}

animal_survey %>%
  group_by(sex) %>%
  summarize(mean_weight = mean(weight, na.rm = TRUE))
```

```{r}
#calling mean on vector of NA values (weights were NA) generating NaN in the first few rows
animal_survey %>%
  group_by(sex, species_id) %>%
  summarize(mean_weight = mean(weight, na.rm = TRUE)) %>% 
  head()

animal_survey %>%
  group_by(sex, species_id) %>%
  summarize(mean_weight = mean(weight, na.rm = TRUE)) %>% 
  tail
```

Remove missing values first by filtering them out and calculating the mean:
```{r}
animal_survey %>%
  filter(!is.na(weight)) %>%
  group_by(sex, species_id) %>%
  summarize(mean_weight = mean(weight))
```

To specify the number of rows displayed, print is used with the number of rows as an argument:
```{r}
animal_survey %>%
  filter(!is.na(weight)) %>%
  group_by(sex, species_id) %>%
  summarize(mean_weight = mean(weight)) %>%
  print(n = 15)
```

To summarize multiple variables simultaneously:
```{r}
animal_survey %>%
  filter(!is.na(weight)) %>%
  group_by(sex, species_id) %>%
  summarize(mean_weight = mean(weight),
            min_weight = min(weight))
```

To rearrange results of the query by minimum weight:
```{r}
animal_survey %>%
  filter(!is.na(weight)) %>%
  group_by(sex, species_id) %>%
  summarize(mean_weight = mean(weight),
            min_weight = min(weight)) %>%
  arrange(min_weight)
```

Sorting results by minimum weight in descending order:
```{r}
animal_survey %>%
  filter(!is.na(weight)) %>%
  group_by(sex, species_id) %>%
  summarize(mean_weight = mean(weight),
            min_weight = min(weight)) %>%
  arrange(desc(mean_weight))
```

### Counting

```{r}
animal_survey %>% 
  count(sex)

#equivalent to count using summarize

animal_survey %>%
    group_by(sex) %>%
    summarise(count = n())

#sorting with count

animal_survey %>%
    count(sex, sort = TRUE) 

#use arrange for better comparison
animal_survey %>%
  count(sex, species) %>%
  arrange(species, desc(n))

```

## Challenge 3

1)
```{r}
animal_survey %>% 
  count(plot_type)
```

2) 
```{r}
animal_survey %>% 
  filter(!is.na(hindfoot_length)) %>%
  group_by(species_id) %>% 
  summarise(mean_hindfoot_len = mean(hindfoot_length),
            min_hindfoot_len = min(hindfoot_length),
            max_hindfoot_len = max(hindfoot_length),
            num_observed = n())
```

3)
```{r}
animal_survey %>% 
  filter(!is.na(weight)) %>%
  group_by(year) %>% 
  filter(weight == max(weight)) %>% 
  select(year, genus, species_id, weight) %>% 
  arrange(year)

```

## Reshaping

### Spread

Transforming animal_survey dataframe to find the mean weight of each genus in each plot over the entire survey period:

```{r}
animal_survey_gw <- animal_survey %>%
  filter(!is.na(weight)) %>%
  group_by(plot_id, genus) %>%
  summarize(mean_weight = mean(weight))

str(animal_survey_gw)
```


```{r}
animal_survey_spread <- animal_survey_gw %>%
  spread(key = genus, value = mean_weight)

str(animal_survey_spread)

animal_survey_gw %>%
  spread(genus, mean_weight, fill = 0) %>%
  head()
```

### Gather

```{r}
animal_survey_gather <- animal_survey_spread %>%
  gather(key = "genus", value = "mean_weight", -plot_id)

str(animal_survey_gather)

#To specify what columns to include 

animal_survey_spread %>%
  gather(key = "genus", value = "mean_weight", Baiomys:Spermophilus) %>%
  head()
```

## Challenge 4

1)
```{r}
num_genera_reshaped <- animal_survey %>%
  group_by(plot_id, year) %>% 
  summarise(num_genera=n_distinct(genus)) %>% 
  spread(key=year, value=num_genera)

num_genera_reshaped
```

2)
```{r}
num_genera_revert <- num_genera_reshaped %>% 
  gather(key="year", value="num_genera", -plot_id)

num_genera_revert
```

3)
```{r}
animal_survey_hindfoot_weight <- animal_survey %>% 
  gather("measurement", "value", hindfoot_length, weight)

animal_survey_hindfoot_weight
```

4)
```{r}
animal_survey_hindfoot_weight %>%
  group_by(year, measurement, plot_type) %>%
  summarize(mean_value = mean(value, na.rm=TRUE)) %>%
  spread(measurement, mean_value)

```

## Exporting Data

Filtering out NA values from observations:
```{r}
animal_survey_complete <- animal_survey %>%
  filter(!is.na(weight),          
         !is.na(hindfoot_length),  
         !is.na(sex))                
```


Extract highly common species ID:
```{r}
species_counts <- animal_survey_complete %>%
    count(species_id) %>% 
    filter(n >= 50)
head(species_counts)
```

Retain the most common species and reassign to animal_survey_complete dataframe:
```{r}

animal_survey_complete <- animal_survey_complete %>%
  filter(species_id %in% species_counts$species_id)

animal_survey_complete
dim(animal_survey_complete)
```

Write data set and keep as CSV file in data folder for future analysis:
```{r}
write_csv(animal_survey_complete, path = "data/surveys_complete.csv")
```

