---
title: "Lab 6"
author: "Vista Sohrab"
date: "9/28/2020"
output: html_document
editor_options: 
  chunk_output_type: console
bibliography: data/references_lab6.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data Maps and Interactive Graphs from the COVID-19 reporting data

Loading the necessary libraries:
```{r load libraries, warning=FALSE}
library(tidyverse)
library(plotly)
library(lubridate)
library(mapdata)
library(maps)
library(viridis)
library(wesanderson)
```

## Exercises

### Exercise 1: 

World COVID-19 report from 09/26/2020.
```{r}
daily_report <- read_csv(url("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_daily_reports/09-26-2020.csv")) %>% 
  rename(Long = "Long_")
```

Summarize the counts for each country and obtain median latitude and longitude from daily_report dataset:

```{r}
deaths_report <- daily_report %>% 
  group_by(Country_Region) %>%
  filter(Lat != "NA") %>% 
  filter(Long != "NA") %>% 
  summarise(Lat = median(Lat), Long = median(Long), Total_Deaths = sum(Deaths))
  
```

World map of COVID-19 deaths reported from 09/26/2020
```{r warning=FALSE}
ggplot(deaths_report, aes(x = Long, y = Lat, size = Total_Deaths/1000)) +
    borders("world", colour = NA, fill = "grey90") +
    theme_bw() +
    geom_point(shape = 21, color='purple', fill='purple', alpha = 0.5) +
    labs(title = 'World COVID-19 Death Report',x = '', y = '',
        size="Deaths (x1000)") +
    theme(legend.position = "right") +
    coord_fixed(ratio=1.5)
```

Interactive world map of deaths reported from 09/26/2020:
```{r}
ggplotly(
ggplot(deaths_report, aes(x = Long, y = Lat, text = Country_Region, size = Total_Deaths/1000)) +
    borders("world", colour = NA, fill = "grey90") +
    theme_bw() +
    geom_point(shape = 21, color='purple', fill='purple', alpha = 0.5) +
    labs(title = 'World COVID-19 Death Report',x = '', y = '',
        size="Deaths (x1000)") +
    theme(legend.position = "right") +
    coord_fixed(ratio=1.5)
)
```

### Exercise 2

Update to Anisa Dhana's graph layout of the uS based on 09/26/2020 data:

```{r}
US_daily_report <- daily_report %>% 
  filter(Country_Region == "US") %>% 
  filter (!Province_State %in% c("Alaska","Hawaii", "American Samoa",
                  "Puerto Rico","Northern Mariana Islands", 
                  "Virgin Islands", "Recovered", "Guam", "Grand Princess",
                  "District of Columbia", "Diamond Princess")) %>% 
  filter(Lat > 0)
  
```


```{r}
mybreaks <- c(1, 100, 1000, 10000, 10000)

ggplot(US_daily_report, aes(x = Long, y = Lat, size = Confirmed)) +
    borders("state", colour = "white", fill = "grey90") +
    geom_point(aes(x=Long, y=Lat, size=Confirmed, color=Confirmed),stroke=F, alpha=0.4)+
    scale_size_continuous(name="Cases", range=c(1,7), trans = "log",
                        breaks=mybreaks, labels = c("1-4,999",
                        "5,000-9,999", "10,000-49,999", "50,000-99,999", "100,000+")) +
    scale_color_viridis_c(option="viridis",name="Cases", trans = "log", 
                         breaks=mybreaks, labels = c("1-4,999",
                        "5,000-9,999", "10,000-49,999", "50,000-99,999", "100,000+")) + 
# Cleaning up the graph
  
  theme_void() + 
    guides(colour = guide_legend()) +
    labs(title = "Anisa Dhana's layout for COVID-19 Confirmed Cases in the US") +
    theme(
      legend.position = "bottom",
      text = element_text(color = "#22211d"),
      plot.background = element_rect(fill = "#ffffff", color = NA), 
      panel.background = element_rect(fill = "#ffffff", color = NA), 
      legend.background = element_rect(fill = "#ffffff", color = NA))+  
    coord_fixed(ratio=1.5)
```

### Exercise 3

Number of Confirmed Cases by US County:
```{r}
library(RColorBrewer)
# To display only colorblind-friendly brewer palettes, specify the option colorblindFriendly = TRUE as follow:
#display.brewer.all(colorblindFriendly = TRUE)


# Get and format the covid report data
US_Sept26_report <- US_daily_report %>% 
  unite(Key, Admin2, Province_State, sep = ".") %>% 
  group_by(Key) %>% 
  summarize(Confirmed = sum(Confirmed)) %>% 
  mutate(Key = tolower(Key))
  
# get and format the map data
us <- map_data("state")
counties <- map_data("county") %>% 
  unite(Key, subregion, region, sep = ".", remove = FALSE)
# Join the 2 tibbles
state_join <- left_join(counties, US_Sept26_report, by = c("Key"))
# sum(is.na(state_join$Confirmed))
ggplot(data = us, mapping = aes(x = long, y = lat, group = group)) + 
  coord_fixed(1.3) + 
  # Add data layer
  borders("state", colour = "black") +
  geom_polygon(data = state_join, aes(fill = Confirmed)) +
  scale_fill_gradientn(colors = brewer.pal(n = 5, name = "GnBu"),
                       breaks = c(1,10, 100, 1000, 10000, 100000),
                       trans = "log10", na.value = "White") +
  ggtitle("Number of Confirmed Cases by US County") +
  theme_bw() 
```

## Exercise 4

Confirmed Cases by County in California:
```{r}
daily_report <- read_csv(url("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_daily_reports/09-26-2020.csv")) %>% 
  filter(Province_State == "California") %>% 
  group_by(Admin2) %>% 
  summarize(Confirmed = sum(Confirmed)) %>% 
  mutate(Admin2 = tolower(Admin2))
us <- map_data("state")
ca_us <- subset(us, region == "california")
counties <- map_data("county")
ca_county <- subset(counties, region == "california")
state_join <- left_join(ca_county, daily_report, by = c("subregion" = "Admin2")) 
# plot state map
ggplot(data = ca_county, mapping = aes(x = long, y = lat, group = group)) + 
  coord_fixed(1.3) + 
# Add data layer
  geom_polygon(data = state_join, aes(fill = Confirmed), color = "white") +
    scale_fill_gradientn(colors = brewer.pal(n = 5, name = "BuPu"), 
                         trans= "log10") + theme_bw()+
  labs(title = "COVID-19 Confirmed Cases in California")
```

Interactive Plot of Confirmed Cases in California:
```{r}
ggplotly(
  ggplot(data = ca_county, mapping = aes(x = long, y = lat, group = group)) + 
  coord_fixed(1.3) + 
# Add data layer
  geom_polygon(data = state_join, aes(fill = Confirmed), color = "black") +
    scale_fill_gradientn(colours = 
                         wes_palette("Moonrise3", 100, type = "continuous")) +
  ggtitle("COVID-19 Cases in CA") +
# Cleaning up the graph
  labs(x=NULL, y=NULL) +
  theme(panel.border = element_blank()) +
  theme(panel.background = element_blank()) +
  theme(axis.ticks = element_blank()) +
  theme(axis.text = element_blank())
)
```

### Exercise 5

This report uses COVID-19 data available from [@Dong2020]. Data analysis is performed in R[@RCoreTeam] by using Tidyverse [@Tidyverse], specifically ggplot for graphs [@ggplot2], and [@plotly] for interactive graphs.

An interactive map of COVID-19 confirmed cases in the US as of 09/26/2020:
```{r echo=FALSE, message=FALSE}
daily_report <- read_csv(url("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_daily_reports/09-26-2020.csv")) %>% 
  filter(Country_Region == "US") %>% 
  group_by(Province_State) %>% 
  summarize(Confirmed = sum(Confirmed)) %>% 
  mutate(Province_State = tolower(Province_State))
# load the US map data
us <- map_data("state")
# We need to join the us map data with our daily report to make one data frame/tibble
state_join <- left_join(us, daily_report, by = c("region" = "Province_State"))
# plot state map
ggplotly(
ggplot(data = us, mapping = aes(x = long, y = lat, group = group)) + 
  coord_fixed(1.3) + 
# Add data layer
  geom_polygon(data = state_join, aes(fill = Confirmed), color = "black") +
  scale_fill_gradientn(colours = 
                         wes_palette("GrandBudapest2", 100, type = "continuous")
                         ) + theme_bw()+
  labs(title = "COVID-19 Confirmed Cases in the US")
)
```

California, Texas, and Florida have the highest number of confirmed cases to date.

It would be informative to view county cases for these states with high confirmed cases. In the case of California, southern and central California seem to be most affected.
```{r echo=FALSE, message=FALSE}
daily_report <- read_csv(url("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_daily_reports/09-26-2020.csv")) %>% 
  filter(Province_State == "California") %>% 
  group_by(Admin2) %>% 
  summarize(Confirmed = sum(Confirmed)) %>% 
  mutate(Admin2 = tolower(Admin2))
us <- map_data("state")
ca_us <- subset(us, region == "california")
counties <- map_data("county")
ca_county <- subset(counties, region == "california")
state_join <- left_join(ca_county, daily_report, by = c("subregion" = "Admin2")) 
# plot state map
ggplotly(
ggplot(data = ca_county, mapping = aes(x = long, y = lat, group = group)) + 
  coord_fixed(1.3) + 
# Add data layer
  geom_polygon(data = state_join, aes(fill = Confirmed), color = "black") +
    scale_fill_gradientn(colors = brewer.pal(n = 5, name = "Purples"), 
                         trans="log10") + theme_bw()+
  labs(title = "COVID-19 Confirmed Cases in California"))
```


For Texas, the cases seem to be much more spread across the state with high number of cases in the south and east of the state.
```{r echo = FALSE, message=FALSE}
daily_report <- read_csv(url("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_daily_reports/09-26-2020.csv")) %>% 
  filter(Province_State == "Texas") %>% 
  group_by(Admin2) %>% 
  summarize(Confirmed = sum(Confirmed)) %>% 
  mutate(Admin2 = tolower(Admin2))
us <- map_data("state")
tx_us <- subset(us, region == "texas")
counties <- map_data("county")
tx_county <- subset(counties, region == "texas")
state_join <- left_join(tx_county, daily_report, by = c("subregion" = "Admin2")) 
# plot state map
ggplotly(
ggplot(data = tx_county, mapping = aes(x = long, y = lat, group = group)) + 
  coord_fixed(1.3) + 
# Add data layer
  geom_polygon(data = state_join, aes(fill = Confirmed), color = "black") +
    scale_fill_gradientn(colors = brewer.pal(n = 5, name = "Purples"), 
                         trans= "log10") + theme_bw()+
  labs(title = "COVID-19 Confirmed Cases in Texas")
)
```

In the case of Florida, counties along the border especially the southern region have the greatest number of confirmed cases.
```{r echo = FALSE, message=FALSE}
daily_report <- read_csv(url("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_daily_reports/09-26-2020.csv")) %>% 
  filter(Province_State == "Florida") %>% 
  group_by(Admin2) %>% 
  summarize(Confirmed = sum(Confirmed)) %>% 
  mutate(Admin2 = tolower(Admin2))
us <- map_data("state")
fl_us <- subset(us, region == "florida")
counties <- map_data("county")
fl_county <- subset(counties, region == "florida")
state_join <- left_join(fl_county, daily_report, by = c("subregion" = "Admin2")) 
# plot state map
ggplotly(
ggplot(data = fl_county, mapping = aes(x = long, y = lat, group = group)) + 
  coord_fixed(1.3) + 
# Add data layer
  geom_polygon(data = state_join, aes(fill = Confirmed), color = "black") +
    scale_fill_gradientn(colors = brewer.pal(n = 5, name = "Purples"), 
                         trans= "log10") + theme_bw()+
  labs(title = "COVID-19 Confirmed Cases in Florida")
)
```

## References